---
title: "Abalone Report - Simple"
author: "Melinda Higgins"
date: "2022-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# knitr options to set up
# show code in report
knitr::opts_chunk$set(echo = TRUE)

# if errors, keep knitting
knitr::opts_chunk$set(error = TRUE)

# hide messages and warnings
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load printr package for nicer
# table output formatting
library(printr)

# load packages needed
library(readr)
library(dplyr)
library(ggplot2)
library(purrr)

# load abalone dataset
abalone <- read_csv("abalone.csv")

# add age variable
abalone <- abalone %>%
  mutate(age = rings + 1.5)
```

## Make a plot of Age (Y-axis) by Whole Weight (X-axis)

Let's make a 2 dimensional scatterplot of the abalone `age` by `wholeWeight`. We'll also overlay a best fit line by `sex.` And we'll add some better labels and title.

Learn more about `ggplot2` graphics at:

* [http://www.cookbook-r.com/Graphs/](http://www.cookbook-r.com/Graphs/)
* [https://r-graphics.org/](https://r-graphics.org/)
* [https://ggplot2-book.org/](https://ggplot2-book.org/)
* and the great examples at [https://r-graph-gallery.com/index.html](https://r-graph-gallery.com/index.html)
* including cutomizing legend titles at [https://r-graph-gallery.com/239-custom-layout-legend-ggplot2.html](https://r-graph-gallery.com/239-custom-layout-legend-ggplot2.html)


Notice that we can SEE the R code used here.

```{r}
ggplot(abalone, aes(wholeWeight, age, color = sex)) +
  geom_point() +
  geom_smooth(method = lm) +
  xlab("Whole Weight in grams") +
  ylab("Age in years (rings + 1.5)") +
  labs(title = "Age of Abalones by Whole Weight",
       subtitle = "Best fit lines shown by sex") +
  labs(color = "Biological Sex")
```

## Inline r code to insert results inside a sentence

Here is an example of writing a sentence with some results that are automatically inserted. For example, the average age of the `r nrow(abalone)` abalones is `r mean(abalone$age)`.

## Making a simple table

Here is a simple table with a caption using the `table()` function and the `knitr::kable()` options for formatting.

```{r}
# a simple table
abalone %>%
  select(sex) %>%
  table()
```

Now adding a caption 

```{r}
# a simple table
abalone %>%
  select(sex) %>%
  table() %>%
  knitr::kable(caption = "Biological Sex of Abalones - Counts")
```

Next add some custom column headings

```{r}
# a simple table
abalone %>%
  select(sex) %>%
  table() %>%
  knitr::kable(caption = "Biological Sex of Abalones - Counts",
               col.names = c("Biological Sex", "Frequency"))
```

## Very brief intro to `gtsummary` package

Another "tidyverse" compatible R package for working with tables is the `gtsummary` package.

Learn more at:

* [https://www.danieldsjoberg.com/gtsummary/index.html](https://www.danieldsjoberg.com/gtsummary/index.html)
* specifically review the vignette for the `tbl_summary()` function [https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html)

The code below will produce a table with summary statistics based on the variable type:

* `sex` is categorical character variable so you get frequencies and percents of the total;
* `age` is continuous numeric variable so you get the median and IQR (interquartile range) by default.

```{r}
library(gtsummary)
abalone %>%
  select(sex, age) %>%
  tbl_summary()
```

Suppose we instead want to get the mean and standard deviation for `age`. Inside the `tbl_summary()` function we can use the `statistic = list(...)` option.

```{r}
abalone %>%
  select(sex, age) %>%
  tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ({sd})")
    )

```

We could take this a step further and get statistics by a grouping variable. For example, let's get the 3 dimensional measurements by `sex`. Inside the `tbl_summary()` function we can use the `by = group` option.

```{r}
abalone %>%
  select(height, diameter, length, sex) %>%
  tbl_summary(
    by = sex,
    statistic = list(all_continuous() ~ "{mean} ({sd})")
    )
```

